Mechanical loading prediction through accelerometry data during walking and running

### 2021-10-01 manuscript/manuscript.Rmd
Add HF suggestions

3 - General Surgery Department, São João Medical Center, Porto, Portugal

@@

Comparing our results with previous studies that used a similar approach to predict pGRF, prediction errors from our models are comparable with some studies whose best results presented a MAPE of approximately 6.0\%, but which included only walking in their protocol [@Fortune2014; @Neugebauer2018; @Veras2020]. Additionally, previous studies that developed pGRF prediction models during both walking and running showed a MAPE of 8.3\% [@Neugebauer2014], which is higher than our results, and 5.2\% [@Neugebauer2012] with no model cross-validation, which is similar to our results when no cross-validation method is applied (highest value of `r accuracy_no_cv`\% in our models). Nevertheless, models from both studies present a caveat, that is the inclusion of a term in their equations that distinguishes whether the activity is walking or running, which is not easy to be identified from free-living accelerometer data, hindering the real-world application of these previous equations. Also, one of these studies included in their sample only children and adolescents [@Neugebauer2012], which present a higher variability in their gait pattern compared to the adults [@Gouelle2016], making this equation unsuitable to use in an adult population. Furthermore, it was observed that the reference equation [@Neugebauer2014] used as reference has a much lower accuracy during walking compared with running (16.6\% vs. 9.7\%). As this reference equation was not validated in a sample including people with obesity, it performs poorly when applied to data from people with high body mass, as is the case in our sample that performed the walking trials, compared with running trials mostly performed by non-obese people. It is also important to mention that the prediction error found in our pGRF models is lower than the error typically found in models developed to estimate energy expenditure [@Diniz-Sousa2020], which is one of the main uses of accelerometer data [@Migueles2017].

@@

Our findings revealed that ankle, lower back or hip were all valid accelerometer placements to accurately predict pGRF. However, despite the similarities, hip was the accelerometer placement with the best prediction accuracy, which reinforces it as the preferable location to collect and analyze accelerometry-measured physical activity data. This placement is already well accepted for energy expenditure prediction and physical activity intensity classification [@Migueles2017] and has also shown a high mechanical loading prediction ability in previous studies [@Veras2020; @Neugebauer2018; @Neugebauer2012]. Besides various placements, the current study also analyzed predictions from both the resultant and vertical vectors. Although ACC and GRF from the vertical vector are the most studied and frequently associated with bone health [@Vainionpaa2006; @Heikkinen2007], the resultant vector of both ACC and GRF during locomotion is greatly determined by the vertical vector, and, as can be observed in Supplemental Figure S1, in this study values from both vectors were very similar. Also, mechanical loading prediction based on the resultant ACC offers the important advantage of not being dependent on the accelerometer positioning and correct orientation, which is hard to assure when measuring daily physical activity [@Welk2005; @Wundersitz2015].

@@

To the best of our knowledge, only two studies have already attempted to develop pLR prediction models based on accelerometry data, and both studies only included walking activities. Accordingly to what we found, these studies also showed that the prediction error for these models is considerably larger than for the pGRF prediction. One of these researches investigated the use of four different accelerometer placements (ankle, tibia, thigh and waist) to predict pLR and achieved a MAPE ranging from 23.4\% to 46.1\% [@Fortune2014], while the other obtained values from 19.5\% to 22.4\% using accelerometers at the lower back and hip [@Veras2020]. Results from the models developed in these previous investigations have both lower and higher prediction errors compared to our results, which were approximately of 35\%. Although pLR is an important variable, known for a long time to determine to a great extent the bone formation response to mechanical loading [@Bassey1998; @Turner1995], based on our findings and in previous studies [@Fortune2014; @Veras2020], its prediction through accelerometer data does not seem to yield trustworthy results.
\medskip

@@

Besides these advantages, our models present some important potential clinical applications. For instance, these models can be used to analyze data from large observational studies and make associations between mechanical loading and bone health. But instead of analyzing accelerometer data averaged in epochs similarly to some previous publications [@Chastin2014; @Stiles2017], which can dilute the higher intensities [@Fabre2020], with our method raw ACC can be analyzed at a higher resolution by identifying the pGRF associated with each step taken and establishing, for example, the number of peaks above a certain magnitude threshold needed to elicit a positive bone adaptation. Another potential application could be the computation of the cumulative mechanical loading by adding up the pGRF of all detected steps [@Diniz-Sousa2021]. This information could be used either for the determination of the relationship between the mechanical loading accumulated in a certain period (e.g., a day) and the bone response to this loading [@Ahola2010], or to investigate the causes of bone loss in populations with bone health issues, such as bariatric patients [@Diniz-Sousa2021]. This approach has already been applied with equations developed in a previous research from our group [@Veras2020], but now can be extended to include both types of locomotion (walking and running) contemplated in the equations from the current study. Another possible application is in sports training, by monitoring the amount of loading beared by athletes and relate this information with osteoarticular safety of the training regimen [@Vanwanseele2020].

@@

Despite the validity of the prediction models developed here, some limitations should be mentioned. First, although the activities included in our protocol (walking and running) are among the most frequently performed motor activities during daily living, our prediction models do not apply to other activities known to influence bone adaptation to mechanical stimuli, such as jumping and direction changing. Second, in the absence of another sample for model validation, the leave-one-out cross-validation method was used, which is the method of choice in these situations [@Staudenmayer2012]. Nevertheless, future studies should test the out-of-sample validity of these models. Moreover, the protocol was performed on an instrumented treadmill. The use of a treadmill, although enhancing the study internal validity by the continuously measuring the GRF [@Ngoh2018] also reduces the external validity due to possible biomechanical changes in the walking and running patterns induced by the treadmill [@Kluitenberg2012]. Lastly, bone adaptation to mechanical stimulation depends not only on external loads, such as GRF, but also on internal loads generated by muscle forces [@Robling2009]. As accelerometers cannot reliably assess muscle force generation, they may have a limited capacity in estimating the bone response to loading in situations where there are no major gravitational loadings involved.

@@

In conclusion, pGRF could be accurately predicted by accelerometry-based regression models. These models are valid for the use of both the resultant vector and its vertical component, and using data from accelerometers placed at either the ankle, lower back or hip. The prediction models developed in this study have proved to be an efficient method to assess pGRF out of laboratory conditions, with potential applications for monitoring the mechanical loading in several contexts and establishing a relationship between mechanical loading and either beneficial or pathological osteoarticular adaptations. This is also the first work, to the best of our knowledge, to explore the prediction of pLR based on accelerometry data during both walking and running, although the accuracy of these models was unsatisfactory.

@@

This study was funded by the Fundação para a Ciência e a Tecnologia (FCT) (grant PTDC/DTP-DES/0968/2014) and by the European Regional Development Fund (ERDF) through the Operational Competitiveness Programme (COMPETE) (grant POCI-01-0145-FEDER-016707). The study was developed in the Research Centre in Physical Activity, Health and Leisure (CIAFEL) funded by ERDF through the COMPETE and by the FCT (grant UIDB/00617/2020), and Laboratory for Integrative and Translational Research in Population Health (ITR). Lucas Veras, Florêncio Diniz-Sousa and Giorjines Boppre were supported by the FCT grants UI/BD/150673/2020, SFRH/BD/117622/2016, SFRH/BD/146976/2019, respectively. The authors thank the participants who took part in this research and all who have collaborated in the project.

>>>>

3 - Centro de Responsabilidade Integrado de Obesidade (CRIO), Centro Hospitalar Universitário de São João, Porto, Portugal

@@

Comparing our results with previous studies that used a similar approach to predict pGRF, prediction errors from our models are comparable with some studies whose best results presented a MAPE of approximately 6.0\%, but which included only walking in their protocol [@Fortune2014; @Neugebauer2018; @Veras2020]. Additionally, previous studies that developed pGRF prediction models during both walking and running showed a MAPE of 8.3\% [@Neugebauer2014], which is higher than our results, and 5.2\% [@Neugebauer2012] with no model cross-validation, which is similar to our results when no cross-validation method is applied (highest value of `r accuracy_no_cv`\% in our models). Nevertheless, models from both studies present a caveat, that is the inclusion of a term in their equations that implies distinguishing whether the activity is walking or running. This is not easy to be identified from free-living accelerometer data, hindering real-world application of these previous equations. Also, one of these studies included only children and adolescents in their sample [@Neugebauer2012], which present a higher variability in terms of gait pattern compared to adults [@Gouelle2016], making, therefore, this equation unsuitable to use in an adult population. Furthermore, it was observed that the equation used as reference in this study [@Neugebauer2014] has a much lower accuracy during walking compared with running (16.6\% vs. 9.7\%). As this reference equation was not validated in a sample including people with obesity, it performs poorly when applied to data from people with high body mass, as is the case in our sample that performed the walking trials, compared with running trials mostly performed by non-obese people. It is also important to mention that the prediction error found in our pGRF models is lower than the error typically found in models developed to estimate energy expenditure [@Diniz-Sousa2020], which is one of the main uses of accelerometer data [@Migueles2017].

@@

Our findings revealed that ankle, lower back or hip were all valid accelerometer placements to accurately predict pGRF. However, despite the similarities, hip was the accelerometer placement with the best prediction accuracy, which reinforces it as the preferable location to collect and analyze accelerometry-measured physical activity data. This placement is already well accepted for energy expenditure prediction and physical activity intensity classification purposes [@Migueles2017] and has also shown a high mechanical loading prediction ability in previous studies [@Veras2020; @Neugebauer2018; @Neugebauer2012]. In addition to testing various accelerometer placements, the current study also investigated the prediction ability of both the resultant and vertical vectors. Although ACC and GRF from the vertical vector are the most studied and frequently associated with bone health variables [@Vainionpaa2006; @Heikkinen2007], the resultant vector of both ACC and GRF during locomotion is greatly determined by the vertical vector. In fact, as can be observed in Supplemental Figure S1, in this study, values from both vectors were very similar. Also, mechanical loading prediction based on the resultant ACC offers the important advantage of not being dependent on the accelerometer positioning and correct orientation, which is hard to assure when measuring daily physical activity [@Welk2005; @Wundersitz2015].

@@

To the best of our knowledge, only two studies have already attempted to develop pLR prediction models based on accelerometry data, and both studies only included walking activities. In accordance with our own findings, these studies also showed that the prediction error for these models is considerably larger than that for the pGRF prediction. One of these studies investigated the use of four different accelerometer placements (ankle, tibia, thigh and waist) to predict pLR and achieved a MAPE ranging from 23.4\% to 46.1\% [@Fortune2014], while the other obtained values from 19.5\% to 22.4\% using accelerometers at the lower back and hip [@Veras2020]. Results from the models developed in these previous studies have both lower and higher prediction errors compared to our own results, which were of approximately 35\%. Although pLR is an important variable, known for a long time to determine to a great extent the bone formation response to mechanical loading [@Bassey1998; @Turner1995], based on our findings and in previous studies [@Fortune2014; @Veras2020], its prediction through accelerometer data, at the current state of technical accelerometer development, does not seem to yield trustworthy results that can be used in models to predict the bone anabolic response to mechanical loading.

@@

In addition to these technical advantages, our models present some important potential clinical applications. For instance, these models can be used to analyze data from large observational studies and to investigate associations between objectively mesured mechanical loading and bone health. Nevertheless, instead of analyzing accelerometer data averaged in epochs as it has been performed previously [@Chastin2014; @Stiles2017], which can have a dilution effect and hinder the detection of higher intensities [@Fabre2020], with our method, raw ACC can be analyzed at a higher resolution by identifying the pGRF associated with each step taken and establishing, for instance, the number of peaks above a certain magnitude threshold needed to elicit a positive bone adaptation. Another potential application could be the computation of the cumulative mechanical loading by adding up the pGRF of all detected steps [@Diniz-Sousa2021]. This information could be used either for the determination of the relationship between mechanical loading accumulated within a certain period (e.g., a day) and the bone response to this loading [@Ahola2010], or to investigate the causes of bone loss in populations with bone health issues, such as post-bariatric surgery patients [@Diniz-Sousa2021]. This approach has already been followed by using equations previously developed by our group [@Veras2020]. However, these new equations described in the current study have the advantage of being able to be applied to both types of locomotion (walking and running) and not just walking. Another possible application is in sports training, by monitoring the amount of mechanical loading sustained by athletes and using this information for training loading monitoring and for establishing adequate measures to prevent osteoarticular overload and reduce the risk of osteoarthritis [@Vanwanseele2020].

@@

Despite the validity of the prediction models developed here, some limitations should be mentioned. First, although the activities included in our protocol (walking and running) are among the most frequently performed motor activities during daily living, our prediction models do not apply to other activities known to influence bone adaptation to mechanical stimuli, such as jumping and direction changing. Second, in the absence of another sample for model validation, the leave-one-out cross-validation method was used, which is the method of choice in these situations [@Staudenmayer2012]. Nevertheless, future studies should test the out-of-sample validity of these models. Moreover, the protocol was performed on an instrumented treadmill. The use of a treadmill, although enhancing the study internal validity by the continuously measuring the GRF [@Ngoh2018] also reduces the external validity due to possible biomechanical changes in the walking and running patterns induced by the treadmill [@Kluitenberg2012]. Lastly, bone adaptation to mechanical stimulation depends not only on external loads, such as GRF, but also on internal loads generated by muscle forces [@Robling2009]. As accelerometers cannot reliably assess muscle force generation, they may have a limited capacity in estimating the bone response to loading in situations where there are no major gravitational loadings involved, such as during resistance training.

@@

In conclusion, pGRF could be accurately predicted by accelerometry-based regression models. These models are valid for the use of both the resultant vector and its vertical component, and for using data derived from accelerometers placed at either the ankle, lower back or hip. The prediction models developed in this study have proved to be an efficient method to assess pGRF out of laboratory conditions, with potential applications for monitoring mechanical loading in several contexts and for establishing a relationship between mechanical loading and either beneficial or pathological skeletal and osteoarticular adaptations. This is also the first work, to the best of our knowledge, to explore the prediction of pLR based on accelerometry data during both walking and running, although the accuracy of these models was found to be unsatisfactory.

@@

This study was funded by the Fundação para a Ciência e a Tecnologia (FCT) (grant PTDC/DTP-DES/0968/2014) and by the European Regional Development Fund (ERDF) through the Operational Competitiveness Programme (COMPETE) (grant POCI-01-0145-FEDER-016707). The study was developed in the Research Centre in Physical Activity, Health and Leisure (CIAFEL) funded by ERDF through the COMPETE and by the FCT (grant UIDB/00617/2020), and Laboratory for Integrative and Translational Research in Population Health (ITR). Lucas Veras, Florêncio Diniz-Sousa and Giorjines Boppre are supported by the FCT grants UI/BD/150673/2020, SFRH/BD/117622/2016 and SFRH/BD/146976/2019, respectively. The authors thank the participants who took part in this research and all who have collaborated in the research project.
